Gradual Soundness: Lessons from Static Python

### 2022-01-05 pj-7.1/paper.tex
pj: sec 3, more model + start describing

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\defeq}{=}
\newcommand{\mfeq}{=}
\newcommand{\langmid}{\mathrel{\mathbf{\Big\vert}}}
\newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
\newcommand{\ipara}[1]{\emph{#1}\/.}

\newcommand{\spann}[2]{#1\!:\!#2}
\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\codefont}[1]{\emph{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\spteval}{\typefont{S}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptbool}{\typefont{Bool}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptdyn}{\typefont{Dynamic}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{None}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
\newcommand{\sptrawpydict}{\typefont{PyDict}}
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}
\newcommand{\sptenv}{\Gamma}
\newcommand{\sptenvnil}{\cdot}
\newcommand{\sptenvempty}{\sptenvnil}
\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
\newcommand{\sptfundef}[3]{#1(#2)\code{ -> }#3}
\newcommand{\sptclassdef}[4]{\mathrm{class}~#1(#2):~#3~#4}

\newcommand{\spx}{\code{x}}
\newcommand{\spf}{\code{f}}
\newcommand{\spc}{\code{c}}
\newcommand{\spprog}{\codefont{prog}}
\newcommand{\spstmt}{\codefont{stmt}}
\newcommand{\spexpr}{\codefont{expr}}
\newcommand{\spvalue}{\codefont{val}}
\newcommand{\vdashsub}[1]{\vdash_{#1}}
\newcommand{\wtprog}{\vdashsub{\mathbf{P}}}
\newcommand{\wtexpr}{\vdashsub{\mathbf{E}}}
\newcommand{\spsubt}{\mathrel{\leq:}}
\newcommand{\spcompat}{\sqsubseteq}
\newcommand{\spmatr}{\prec}
\newcommand{\spvardef}[3]{\sptvardef{#1}{#2} = #3}
\newcommand{\spfundef}[4]{\mathrm{def}~#1(#2)\code{ -> }#3: #4}
\newcommand{\spclassdef}[4]{\mathrm{class}~#1(#2):~#3~#4}
\newcommand{\spself}{\code{self}}
\newcommand{\spobject}{\code{object}}
\newcommand{\spnone}{\code{none}}
\newcommand{\spint}{\codefont{int}}
\newcommand{\spbool}{\codefont{bool}}
\newcommand{\spfloat}{\codefont{float}}
\newcommand{\sppydict}[1]{\code{\{}#1\code{\}}}
\newcommand{\spchkdict}[1]{\code{chkdict}(#1)}
\newcommand{\spapp}[2]{#1(#2)}
\newcommand{\spref}[2]{#1[#2]}
\newcommand{\spset}[3]{#1[#2] = #3}

@@

\subsection{Runtime System Highlights}
%% Non-Highlights:
%% - shadow frames = JIT feature, lite array to track Python call stack
%% - strict modules ... avoids toplevel function wrapper

@@

@@

\section{Model of Core \SP{}}
\label{s:model}
%% purpose = types, checks, and optimizations can be formalized and analyzed

@@

%% TODO 2021-12-30 outline
%% NOTE strict modules not modeled
%% - what is language
%% - what properties
%% - what did in model, how to navigate?
%% - bug reports

@@

%% TODO 2021-12-28
%% - model soundness argument = similar to Nom, Redex tests
%%   - Nom has dispatch modes .. SP does not, simpler
%%     - are dyn -> stat overrides allowed and do they affect soundness?
%% - while True break another example of modeling helping out
%%   it's a second implementation, helpful to find bugs
%%   TODO if SP folks say that "continue" version matters for production, ... umm something to think about for us

\subsection{Goals and Limitations}

Our model of \SP{} includes the key aspects of its types,
statements, and values.
The model has two main goals:
illustrate the boundaries where typed and untyped code can mix,
and show how to soundly protect these boundaries.
Compared to other gradual languages, these boundaries are quite restrictive---which
partly explains \SP{} performance~(\cref{s:eval}).
Indeed, the model is nearly a subset of the model for Nom~\cite{mt-oopsla-2021},
hence we focus on explaining the gaps and omit formal proofs.

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle soundly (FILL examples ... while loop?).
Omitting these lets us focus on the difficult corners.
Other Python features are left out because their \SP{} semantics is identical
to Python.
These include \code{eval}, first-class classes, multiple inheritance, and module-level cells.
\SP{} does not rely on types for these features to guide program transformations;
see \cref{s:impl} for further discussion.

%% TODO model omits primitives because their typing is easy, separate hierarchy, clear boundaries
%% TODO paper-model focus on boundaries ... where can dyn slip in ... not a full semantics ...

@@

\begin{figure}[t]
  \begin{langarray}
    \spprog & \defeq &
      \spstmt \langmid \spstmt,~\spprog
  \\
    \spstmt & \defeq &
      \spvardef{\spx}{\sptype}{\spexpr} \langmid
  \\ & &
      \spfundef{\spf}{\spann{\spx}{\sptype}}{\sptype}{\spexpr} \langmid
  \\ & &
      \spclassdef{\spc}{\spc}{\spvardef{\spx}{\sptype}{\spexpr} \ldots}{\spfundef{\spf}{\spself, \spann{\spx}{\sptype}}{\sptype}{\spexpr} \ldots}
  \\
    \spexpr & \defeq &
      \spx \langmid \spnone \langmid \spobject \langmid \spint \langmid \spbool \langmid \spfloat \langmid
  \\ & &
      \sppydict{\spx: \spexpr, \ldots} \langmid
      \spchkdict{\sppydict{\spx: \spexpr, \ldots}} \langmid
  \\ & &
      \spapp{\spx}{\spexpr} \langmid
      \spref{\spexpr}{\spexpr} \langmid
      \spset{\spexpr}{\spexpr}{\spexpr}
  \\
    \sptype & \defeq &
      \sptdyn \langmid
      \sptnone \langmid
      \sptclass \langmid
      \sptobject \langmid
      \sptint \langmid
      \sptbool \langmid
      \sptfloat \langmid
  \\ & &
      \sptpydict{\sptype}{\sptype} \langmid
      \sptchkdict{\sptype}{\sptype} \langmid
      \sptunion{\sptype, \ldots}
  \\
    \sptenv & \defeq &
      \sptenvnil \langmid
      \sptvardef{\spx}{\sptype},~\sptenv \langmid
      \sptfundef{\spf}{\sptype}{\sptype},~\sptenv \langmid
      \sptclassdef{\spc}{\spc}{\sptvardef{\spx}{\sptype} \ldots}{\sptfundef{\spf}{\sptclass, \sptype}{\sptype} \ldots},~\sptenv
  \\
    \spx, \spf, \spc & \defeq & \mbox{variable names}
  \end{langarray}

@@

@@

\begin{figure}[t]
  \fbox{$\sptenv \wtprog \spprog \dashv \sptenv$}
  \begin{mathpar}
    \inferrule*{
      \sptenv_0 \wtprog \spstmt_0 \dashv \sptenv_1
      \\\\
      \sptenv_1 \wtprog \spprog_0 \dashv \sptenv_2
    }{
      \sptenv_0 \wtprog \spstmt_0, \spprog_0 \dashv \sptenv_2
    }
  \end{mathpar}

  \fbox{$\sptenv \wtprog \spstmt \dashv \sptenv$}
  \begin{mathpar}
    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
    }{
      \sptenv_0 \wtprog \spvardef{\spx_0}{\sptype_0}{\spexpr_0} \dashv \sptvardef{\spx_0}{\sptype_0},~\sptenv_0
    }

@@

    \inferrule*{
      \sptenv_1 = \sptclassdef{\spc_0}{\spc_1}{\sptvardef{\spx_0}{\sptype_0} \ldots}{\sptfundef{\spf_n}{\spc_0, \sptype_{dn}}{\sptype_{cn}}}
      \\\\
      %% TODO
      \emph{lookup parent}
      \\\\
      \emph{overrides are subtypes (not merely compatible)}
      \\\\
      \emph{fields and methods are well typed}
    }{
      \sptenv_0 \wtprog \spclassdef{\spc_0}{\spc_1}{\spvardef{\spx_0}{\sptype_0}{\spexpr_0} \ldots}{\spfundef{\spf_n}{\spself, \spann{\spx_n}{\sptype_{dn}}}{\sptype_{cn}}{\spexpr_n}} \dashv \sptenv_1
    }
  \end{mathpar}

  \fbox{$\sptenv \wtexpr \spexpr : \sptype$} (selected rules)
  \begin{mathpar}
    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_1
      \\
      \sptype_1 \spcompat \sptype_0
    }{
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
    }

@@

    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
      \\
      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
    }{
      \sptenv_0 \wtexpr \spexpr_0[\spexpr_1] : \sptdyn
    }

    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\sptype_0}{\sptype_1}
      \\
      \sptenv_0 \wtexpr \spexpr_1 : \sptype_0
    }{
      \sptenv_0 \wtexpr \spexpr_0[\spexpr_1] : \sptype_1
    }

@@

  \fbox{$\sptype \spsubt \sptype$}
  \begin{mathpar}
    \inferrule*{
    }{
      \sptype_0 \spsubt \sptobject
    }

@@

    \inferrule*{
    }{
      \sptype_0 \spsubt \sptype_0
    }

    \inferrule*{
      \sptype_0 \spsubt \sptype_1
      \\
      \sptype_1 \spsubt \sptype_2
    }{
      \sptype_0 \spsubt \sptype_2
    }
  \end{mathpar}

@@

  \(\begin{array}{l@{\quad}l}
    \mbox{Eval. type $\spteval$} & \mbox{Description of run-time check} \\\hline
    \sptoptional{\spteval_0} & \mbox{Accepts either the none value or values that match $\spteval_0$} \\
    \sptrawpydict & \mbox{Accepts any Python dictionary} \\
    \sptchkdict{\spteval_0}{\spteval_1} & \mbox{Accepts checked dictionaries with the same key and value types (${\spteval_0, \spteval_1}$)} \\
    \sptdyn & \mbox{No check needed, accepts any value} \\
    \sptobject & \mbox{Accepts any non-primitive value} \\  %% except primitives
    \sptclass_0 & \mbox{Accepts instances of class $\sptclass_0$} \\
    \sptnone & \mbox{Accepts the Python none value} \\
    \sptint & \mbox{Accepts integers} \\
    \sptbool & \mbox{Accepts booleans (\code{0} and \code{1})} \\
    \sptfloat & \mbox{Accepts floats}
  \end{array}\)
\end{table}

@@

\begin{center}
  %% most filed by KC, a few by Ben
  \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/LuKC1024}
\end{center}

@@

>>>>

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\defeq}{=}
\newcommand{\mfeq}{=}
\newcommand{\langmid}{\mid} %% {\mathrel{\mathbf{\Big\vert}}}
\newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
\newcommand{\ipara}[1]{\emph{#1}\/.}

\newcommand{\spapp}[2]{#1\,(#2)}
\newcommand{\spann}[2]{#1\!:\!#2}
\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\codefont}[1]{\emph{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\spteval}{\typefont{S}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptbool}{\typefont{Bool}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptdyn}{\typefont{Dynamic}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{None}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
\newcommand{\sptrawpydict}{\typefont{PyDict}}
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}
\newcommand{\sptenv}{\Gamma}
\newcommand{\sptenvnil}{\cdot}
\newcommand{\sptenvempty}{\sptenvnil}
\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
\newcommand{\sptfundef}[3]{\spapp{#1}{#2}\code{ -> }#3}
\newcommand{\sptclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}

\newcommand{\spx}{\code{x}}
\newcommand{\spf}{\code{f}}
\newcommand{\spc}{\code{c}}
\newcommand{\spprog}{\codefont{prog}}
\newcommand{\spstmt}{\codefont{stmt}}
\newcommand{\spexpr}{\codefont{expr}}
\newcommand{\spvalue}{\codefont{val}}
\newcommand{\vdashsub}[1]{\vdash_{#1}}
\newcommand{\wtprog}{\vdashsub{\mathbf{P}}}
\newcommand{\wtexpr}{\vdashsub{\mathbf{E}}}
\newcommand{\spsubt}{\mathrel{<:}}
\newcommand{\spsubteq}{\mathrel{\leq:}}
\newcommand{\spcompat}{\sqsubseteq}
\newcommand{\spmatr}{\prec}
\newcommand{\spvardef}[3]{\sptvardef{#1}{#2} = #3}
\newcommand{\spfundef}[4]{\mathrm{def}~\spapp{#1}{#2}\code{ -> }#3: #4}
\newcommand{\spclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\spself}{\code{self}}
\newcommand{\spobject}{\code{object}}
\newcommand{\spnone}{\code{none}}
\newcommand{\spint}{\codefont{int}}
\newcommand{\spbool}{\codefont{bool}}
\newcommand{\spfloat}{\codefont{float}}
\newcommand{\sppydict}[1]{\code{\{}#1\code{\}}}
\newcommand{\spchkdict}[1]{\spapp{\code{chkdict}}{#1}}
\newcommand{\spenvapp}[2]{\spapp{#1}{#2}}
\newcommand{\spdictref}[2]{#1[#2]}
\newcommand{\spdictset}[3]{\spdictref{#1}{#2} = #3}
\newcommand{\spobjref}[2]{#1.#2}
\newcommand{\spobjset}[3]{\spobjref{#1}{#2} = #3}
\newcommand{\spobjapp}[3]{\spobjref{#1}{\spapp{#2}{#3}}}

@@

\subsection{Runtime System Highlights}
%% Non-Highlights:
%% - shadow frames = JIT feature, lite array to track Python call stack

@@

%% TODO move prim.types up?

@@

\subsubsection{Strict Modules}
\label{s:strict-mod}

Python modules are mutable collections of names.
Names can be added, deleted, and updated at runtime.
\SP{} accommodates these behaviors in the same manner as standard Python, which
necessarily limits the benefits of module types.
As an alternative, \SP{} programmers may opt in to a \emph{strict module} semantics.
Strict modules are immutable, and thus enable type-based compiler optimizations.

\section{Formal Model}
\label{s:model}
%% purpose = types, checks, and optimizations can be formalized and analyzed

@@

@@

%% TODO 2021-12-28
%% - while True break another example of modeling helping out
%%   it's a second implementation, helpful to find bugs

To validate the design of \SP{}, we have developed a mechanized model using PLT
Redex~\cite{RYR}.
The model encompasses two soundness claims:
%% addresses two soundness concerns?
\begin{enumerate}
  \item
    \SP{}'s overall approach to gradual interactions is sound; namely, that interactions
    with untyped code do not undermine the validity of types.
  \item
    The typed half of \SP{} is a sound type system for Python.
\end{enumerate}
This section explores the first claim in depth by way of a tiny formal model
that illustrates the \SP{} gradual interaction strategy.
The second claim is justified by our Redex development;
in particular, by its test suite.
This section describes the Redex development in broad strokes.

FILL pick names to avoid confusion between the Redex model and the tiny paper model?

\subsection{Goals and Non-Goals}

Our Redex model of \SP{} includes the key aspects of its types,
statements, and values.
The model has two main goals:
illustrate the boundaries where typed and dynamic code can mix,
and show how to soundly protect these boundaries.
Compared to other gradual languages, these boundaries are quite restrictive---which
partly explains how it is able to perform well in practice~(\cref{s:eval}).
Indeed the boundaries are effectively a subset of those in Nom~\cite{mt-oopsla-2021},
which is a compelling argument for their soundness.

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle soundly (e.g. primitive types~\cref{s:c-types} and strict modules~\cref{s:strict-mod}).
Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
are also omitted.
Other Python features are left out because their \SP{} semantics is identical
to the baseline Python semantics.
These features include \code{eval}, first-class classes, multiple inheritance, and module-level cells;
see \cref{s:impl} for further discussion.

Lastly, the model does not include type-directed optimizations.
Validating the correctness of optimizations is a separate task that requires
significant components; namely, a model of the \SP{} bytecode and a faithful
rendering of its type-based transformations.
Our model establishes soundness as a baseline.
Future work can use soundness to reason informally about optimizations,
or even take it as a starting point for compiler verification.

@@

\begin{figure}[t]
  \begin{langarray}
    \spprog & \defeq &
      \spstmt \langmid \spstmt,~\spprog
  \\
    \spstmt & \defeq &
      \spvardef{\spx}{\sptype}{\spexpr} \langmid
  \\ & &
      \spfundef{\spf}{\spann{\spx}{\sptype}}{\sptype}{\spexpr} \langmid
  \\ & &
      \spclassdef{\spc}{\spc}{\spvardef{\spx}{\sptype}{\spexpr}}{\spfundef{\spf}{\spself, \spann{\spx}{\sptype}}{\sptype}{\spexpr}}
  \\
    \spexpr & \defeq &
      \spx \langmid \spnone \langmid \spobject \langmid \spint \langmid \spbool \langmid \spfloat \langmid
      \sppydict{\spx: \spexpr, \ldots} \langmid
      \spchkdict{\sppydict{\spx: \spexpr, \ldots}} \langmid
  \\ & &
      \spapp{\spx}{\spexpr} \langmid
      \spdictref{\spexpr}{\spexpr} \langmid
      \spdictset{\spexpr}{\spexpr}{\spexpr} \langmid
      \spobjref{\spc}{\spx} \langmid
      \spobjset{\spc}{\spx}{\spexpr} \langmid
      \spobjapp{\spc}{\spf}{\spexpr}
  \\
    \sptype & \defeq &
      \sptdyn \langmid
      \sptnone \langmid
      \sptclass \langmid
      \sptobject \langmid
      \sptint \langmid
      \sptbool \langmid
      \sptfloat \langmid
  \\ & &
      \sptpydict{\sptype}{\sptype} \langmid
      \sptchkdict{\sptype}{\sptype} \langmid
      \sptunion{\sptype, \ldots}
  \\
    \sptenv & \defeq &
      \sptenvnil \langmid
      \sptvardef{\spx}{\sptype},~\sptenv \langmid
      \sptfundef{\spf}{\sptype}{\sptype},~\sptenv \langmid
      \sptclassdef{\spc}{\spc}{\sptvardef{\spx}{\sptype}}{\sptfundef{\spf}{\sptclass, \sptype}{\sptype}},~\sptenv
  \\
    \spx, \spf, \spc & \defeq & \mbox{variable names}
  \end{langarray}

@@

The \SP{} surface syntax is the same as Python 3.8.
Programmers may write type annotations using the PEP 484 syntax;
unannotated code gets the dynamic type by default.
The available static types provide a nominal type system with first-order functions.
These surface types are meant to be descriptive.
By design, they are not fully sound.

\Cref{f:surface-types} illustrates the key aspects of the \SP{} syntax and types
using a tiny grammar.
A program is a sequence of statements;
a statement defines a variable, function, or class.
These definitions may only appear on the top level
and they all require full type annotations.
Functions must have one positional argument.
Classes must declare one parent (perhaps $\spobject$), one field, and one method.
Expressions describe values and simple computations.
The basic values consist of the following: the none value, the top object, integers,
booleans (technically the integers \code{0} and \code{1}), and floating point numbers.
There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
The remaining forms express function calls, dictionary reads and writes, object field reads and writes,
and method calls.
Types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
user-defined class, and union types.
There is no function type; the only way to define a higher-order function is to use the dynamic type.

Relative to \SP{} and our Redex model, the formalization in \cref{f:surface-types} omits
many details of Python including imports, conditionals, and exception handlers.
These details are crucial in the context of the Redex model, which tests whether \SP{}
soundly approximates Python.
They are less important here, where our focus is on type boundaries.
None of the omitted features give substantially new ways for typed code to interact with dynamic code.

@@

\begin{figure}[t]
  \fbox{$\sptenv \wtprog \spprog \dashv \sptenv$}
  \begin{mathpar}
    \inferrule*{
      \sptenv_0 \wtprog \spstmt_0 \dashv \sptenv_1
      \\\\
      \sptenv_1 \wtprog \spprog_0 \dashv \sptenv_2
    }{
      \sptenv_0 \wtprog \spstmt_0, \spprog_0 \dashv \sptenv_2
    }

    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
    }{
      \sptenv_0 \wtprog \spvardef{\spx_0}{\sptype_0}{\spexpr_0} \dashv \sptvardef{\spx_0}{\sptype_0},~\sptenv_0
    }

@@

    \inferrule*{
      \sptenv_1 = \sptclassdef{\spc_0}{\spc_1}{\sptvardef{\spx_0}{\sptype_0}}{\sptfundef{\spf_1}{\spc_0, \sptype_1}{\sptype_2}}, \sptenv_0
      \\\\
      %% TODO are field overrides exact?
      \mbox{if $\spann{\spx_0}{\sptype_0'} \in \spenvapp{\sptenv_1}{\spc_1}$ then $\sptype_0 = \sptype_0'$}
      \\\\
      %% TODO can T override dyn?
      \mbox{if $\sptfundef{\spf_1}{\spc_2, \sptype_1'}{\sptype_2'} \in \spenvapp{\sptenv_1}{\spc_1}$ then $\sptype_1' \spsubteq \sptype_1$ and $\sptype_2 \spsubteq \sptype_2'$}
      \\\\
      \sptenv_1 \wtexpr \spvardef{\spexpr_0} : \sptype_0
      \\
      \spann{\spself}{\spc_0}, \spann{\spx_1}{\sptype_1}, \sptenv_1 \wtexpr \spexpr_1 : \sptype_2
    }{
      \sptenv_0 \wtprog \spclassdef{\spc_0}{\spc_1}{\spvardef{\spx_0}{\sptype_0}{\spexpr_0}}{\spfundef{\spf_1}{\spself, \spann{\spx_1}{\sptype_1}}{\sptype_2}{\spexpr_1}} \dashv \sptenv_1
    }
  \end{mathpar}

  \fbox{$\sptenv \wtexpr \spexpr : \sptype$}~selected rules
  \begin{mathpar}
    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_1
      \\
      \sptype_1 \spcompat \sptype_0
    }{
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
    }

@@

    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
      \\\\
      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
    }{
      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptdyn
    }

    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\sptype_0}{\sptype_1}
      \\\\
      \sptenv_0 \wtexpr \spexpr_1 : \sptype_0
    }{
      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptnone
    }

    \inferrule*{
      \spann{\spx_0}{\sptype_0} \in \spenvapp{\sptenv_0}{\spc_0}
    }{
      \sptenv_0 \wtexpr \spobjref{\spc_0}{\spx_0} : \sptype_0
    }

    \inferrule*{
      \spann{\spx_0}{\sptype_0} \in \spenvapp{\sptenv_0}{\spc_0}
      \\\\
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
    }{
      \sptenv_0 \wtexpr \spobjset{\spc_0}{\spx_0}{\spexpr_0} : \sptnone
    }

    \inferrule*{
      \sptfundef{\spf_0}{\spc_0, \sptype_0}{\sptype_1} \in \spenvapp{\sptenv_0}{\spc_0}
      \\\\
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
    }{
      \sptenv_0 \wtexpr \spobjapp{\spc_0}{\spf_0}{\spexpr_0} : \sptype_1
    }

@@

  \fbox{$\sptype \spsubteq \sptype$} reflexive, transitive closure of $\spsubt$, defined below:
  \begin{mathpar}
    \inferrule*{
    }{
      \sptype_0 \spsubt \sptobject
    }

@@

%    \inferrule*{
%    }{
%      \sptype_0 \spsubt \sptype_0
%    }
%
%    \inferrule*{
%      \sptype_0 \spsubt \sptype_1
%      \\
%      \sptype_1 \spsubt \sptype_2
%    }{
%      \sptype_0 \spsubt \sptype_2
%    }
  \end{mathpar}

@@

  \(\begin{array}{l@{\quad}l}
    \mbox{Eval. Type $\spteval$} & \mbox{Description of run-time check} \\\hline
    \sptoptional{\spteval_0} & \mbox{Accepts either the none value or values that match $\spteval_0$} \\
    \sptrawpydict & \mbox{Accepts any Python dictionary} \\
    \sptchkdict{\spteval_0}{\spteval_1} & \mbox{Accepts checked dictionaries with the same key and value types (${\spteval_0, \spteval_1}$)} \\
    \sptdyn & \mbox{No check needed, accepts any value} \\
    \sptobject & \mbox{Accepts any non-primitive value} \\  %% except primitives
    \sptclass_0 & \mbox{Accepts instances of class $\sptclass_0$} \\
    \sptnone & \mbox{Accepts the Python none value} \\
    \sptint & \mbox{Accepts integers} \\
    \sptbool & \mbox{Accepts booleans (\code{0} and \code{1})} \\
    \sptfloat & \mbox{Accepts floats}
  \end{array}\)
\end{table}

@@

\begin{center}
  %% most filed by KC, a few by Ben
  %% + Carl https://github.com/facebookincubator/cinder/issues/64
  \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/LuKC1024}
\end{center}

@@

%% 2021-01-04 TODO have no evidence that "any module" typing works out ... team
%% tried hard to get network effects ... "We explicitly aimed for converting
%% related modules like this to improve network effects and get better
%% performance."
